---
title: "UDPipe Natural Language Processing - Try it out"
author: "Jan Wijffels"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: false
    toc: false
vignette: >
  %\VignetteIndexEntry{UDPipe Natural Language Processing - Try it out}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, cache=FALSE}
options(width = 1000)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = NA, eval = TRUE)
```

Install the R package. 

```{eval=FALSE}
install.packages("udpipe")
```

## Example

Get your language model and start annotating.

```{r, results='hide'}
library(udpipe)
udmodel <- udpipe_download_model(language = "dutch")
udmodel <- udpipe_load_model(file = udmodel$file_model)
x <- udpipe_annotate(udmodel, 
                     x = "Ik ging op reis en ik nam mee: mijn laptop, mijn zonnebril en goed humeur.")
x <- as.data.frame(x)
```

The annotation returns paragraphs, sentences, tokens, morphology elements like the lemma, the universal part of speech tag and the treebank-specific parts of speech tag, morphosyntactic features and returns as well the dependency relationship. More information at http://universaldependencies.org/guidelines.html

```
str(x)
```

## A small note on encodings

Mark that it is important that the `x` argument to `udpipe_annotate` is in UTF-8 encoding. You can check the encoding of your text with `Encoding('your text')` and to convert your text to UTF-8, use standard R utilities to convert to the right encoding 
as in `iconv('your text', from = 'latin1', to = 'UTF-8')` where you replace the from part with whichever encoding you have your text in, possible your computers default as defined in `localeToCharset()`. So annotation would look something like this: `udpipe_annotate(udmodel, x = iconv('your text', from = 'latin1', to = 'UTF-8'))` or like this `udpipe_annotate(udmodel, x = iconv('your text', to = 'UTF-8'))` if your text is in the encoding of the current locale of your computer.